The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-03-05 paper.tex
Line numbers
Structure editors, some renaming

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\hideLIPIcs

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\kvd}[1]{\textsf{\bfseries #1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\suggest}{\textnormal{\ident{suggest}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
\newcommand{\preview}{\textnormal{\ident{preview}}}
\newcommand{\links}{\textnormal{\ident{links}}}
\DeclareMathOperator{\argmax}{arg\,max}

@@

\author{Tomas Petricek}{Charles University, Prague, Czechia}{tomas@tomasp.net}{0000-0002-7242-2208}{}
\author{Jan Liam Verter}{Charles University, Prague, Czechia}{todo@todo}{}{}
\author{Mikoláš Fromm}{Charles University, Prague, Czechia}{todo@todo}{}{}
\ccsdesc[500]{Software and its engineering~General programming languages}
\keywords{Interactive programming systems, Type providers, Proof assistants}
\nolinenumbers

@@

@@

The definition is not a programming language calculus in the usual sense in that it does not
define a concrete syntax with reduction rules. It is an abstract algebraic structure that captures
the structure of a system that supports the choose-your-own-adventure interaction pattern.
The definition is close to that of an AI assistant \cite{petricek-2023-aias}, which is written
using a language specific for the data wrangling domain (such as cleaning scripts or input and
output data) but is structurally similar. It is also worth noting that the definition may describe
not just trees, but also graphs with cycles -- a system can return to an already visited state.
This is not practically useful, but it does not pose a theoretical problem.

\subparagraph{External mode of embedding.}
One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
concerns the different ways in which a trace of the interaction is embedded in the interactively
constructed program. In the \emph{external mode}, the interaction results in code that becomes a
part of the edited program, but it is not possible to reconstruct the steps used to generate the code.

The choose-your-own-adventure interaction pattern is typically used to complete a partial program.
To model this, we assume that the host language has a notion of a hole, written as~$?$
and that a user can select a part of program to invoke the completion on. We write $E[e]$ for a
completion context, akin to evaluation contexts in operational semantics.

We assume that, for a program containing a hole in a completion context $E[?]$, we can construct
an initial choose-your-own-adventure state using an operation
$\ident{init}(E[?])=\sigmaN$.

\begin{definition}\label{def:external}
An expression $E[?]$ is completed as $E[e]$ via external embedding of an interaction with
a choose-your-own-adventure system consisting of \choices\ and \select\ if:

@@

As we will see when we revisit the earlier examples formally, the external mode of embedding is
used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
In those systems, the user triggers the completion on a proof (program) containing a hole.
They then fill the hole and, possibly iteratively, further holes in the generated proof. The final
expression is embedded in the source code, but it does not indicate what options, identified by
$\iota_1, \ldots, \iota_n$, were selected in the process.

\subparagraph{Internal mode of embedding.}
In the \emph{internal mode}, a trace of the interaction with a choose-your-own-adventure system
is embedded directly in the constructed program. This is the case with type providers, where a
user chooses a sequence of object members to be accessed. The same would be the case in a completion
system for Coq that would offer tactics to apply, becuase the resulting proof would contain a
record of the selected tactics.

To talk about the internal mode formally, we again need the \ident{init} operation, but also an
operation \ident{decode} that extracts identifiers of invoked completions from an expression.
An internal embedding is the same as external embedding with an additional constraint:

\begin{definition}\label{def:internal}
An expression $E[?]$ is completed as $E[e]$ via internal embedding of an interaction with
a choose-your-own-adventure system consisting of \choices\ and \select\ if:

\vspace{-0.5em}
\raggedright
\begin{enumerate}
\item $E[?]$ is completed as $E[e]$ via external embedding using \ident{init}, \choices\ and \select\\
  through a series of choices designated by identifiers $\iota_1, \ldots, \iota_n$,
\item it also holds that $\ident{decode}(e)=(\iota_1, \ldots, \iota_n)$.
\end{enumerate}
\end{definition}

If a choose-your-own-adventure system is integrated in a programming language
through internal embedding, we can reconstruct the choices through which the user constructed
an expression $e$ in a completion context $E[e]$, assuming they used the interactive system rather
than entering the code directly. This also means that we can reconstruct the final state $\sigma_n$
of the system by starting from $\ident{init}(E[?])$ and following the choices specified by
$\iota_1, \ldots, \iota_n$.

@@

@@

\item For AI assistants (Section~\ref{sec:examples-aias}), the system offers a set of constraints
  based on the current inferred program. It does not let the user construct arbitrary constraints.
  Moreover, because the $\mathit{arg\,max}$ operation used in $\select$ performs statistical
  optimization, there is no guarantee that it can be used to generate a specific program.
  The system is only complete if it is possible to choose a constraint set $\boldsymbol{c}$ that
  restrict the set of programs $E_\boldsymbol{c}$ to a single given program. This is the case
  only for some AI assistants (e.g., datadiff, which always offers constraints to map column to
  any chosen other column).

@@

@@

\section{Limitations}

@@

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\hideLIPIcs
\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\kvd}[1]{\textsf{\bfseries #1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\suggest}{\textnormal{\ident{suggest}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
\newcommand{\preview}{\textnormal{\ident{preview}}}
\newcommand{\links}{\textnormal{\ident{links}}}
\DeclareMathOperator{\argmax}{arg\,max}

@@

\author{Tomas Petricek}{Charles University, Prague, Czechia}{tomas@tomasp.net}{0000-0002-7242-2208}{}
\author{Jan Liam Verter}{Charles University, Prague, Czechia}{todo@todo}{}{}
\author{Mikoláš Fromm}{Charles University, Prague, Czechia}{todo@todo}{}{}
\ccsdesc[500]{Software and its engineering~General programming languages}
\keywords{Interactive programming systems, Type providers, Proof assistants}
%\nolinenumbers

@@

\begin{figure}[t]
  \includegraphics[width=0.55\textwidth]{fig/sandblocks.png}
  \caption{Constructing a program using the Sandblocks structure editor \cite{beckmann-2023-all}.
    The user is typing the start of an expression and context menu offers possible production
    rules according to a grammar.}
  \label{fig:sand}
\end{figure}

\subparagraph{Structure editors.}
Structure editors make it possible to construct programs by manipulating the abstract syntax tree
of a program rather than by working with textual representation. The interaction with a structure
editor can be based on menus \cite{teitelbaum-1981-synth}, key bindings \cite{weber-2017-editors},
tiles~\cite{moon-2022-tylr} and a range of other approaches.

Sandblocks~\cite{beckmann-2023-all} is an example of a recent structure editor that is interesting
for multiple reasons. First, the editor is automatically generated from a grammar. Second, the
editor combines keyboard input with context menus (Figure~\ref{fig:sand}). In the above example,
the user is typing code and needs to fill a placeholder for an expression. After they type the
start of the expression, the editor offers a completion list with possible production rules from
the grammar that are valid in the given context and are compatible with the text typed so far. The
user can continue typing (to further refine the recommendation), but they can also select a rule
from the menu. This completes the expression according to the production rule, leaving further
empty holes in the code.

In Sandblocks, the context menus serve more as hints and users of the editor often
construct programs through typing, but the system illustrates the fact that structure editors
can also be based on the interaction pattern captured in this paper. The editor can offer
all possible production rules based on the given grammar and the user could construct an
entire program by choosing one of the rules (and manually entering primitive terminals such as
strings and identifiers).

@@

The definition is not a programming language calculus in the usual sense in that it does not
define a concrete syntax with reduction rules. It is an abstract algebraic structure that captures
the structure of a system that supports the choose-your-own-adventure interaction pattern.
The definition is close to that of an AI assistant \cite{petricek-2023-aias}, which is written
using a language specific for the data wrangling domain (such as cleaning scripts or input and
output data) but is structurally similar. It is also worth noting that the definition may describe
not just trees, but also graphs with cycles -- a system can return to a previously visited state.
This is not practically useful, but it does not pose a theoretical problem.

\subparagraph{Expression completion.}
One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
is whether a trace of the interaction can be reconstructed from the source code of an interactively
constructed program. That is, if we follow a sequence of choices $\iota_1, \ldots, \iota_n$ to
construct a program $e$, is it possible to recover the original sequene of choices from the
program $e$.

The choose-your-own-adventure interaction pattern is typically used to complete a partial program.
To model this, we assume that the host language has a notion of a hole, written as~$?$
and that a user can select a part of program to invoke the completion on. When invoked on
an expression containing multiple holes, the system can start by completing the first hole.

We write $E[e]$ for a completion context, akin to evaluation contexts in operational semantics.
We assume that, for a program containing a hole in a completion context $E[?]$, we can construct
an initial choose-your-own-adventure state using an operation $\ident{init}(E[?])=\sigmaN$.

\begin{definition}[Expression completion]
An expression $E[?]$ is completed as $E[e]$ through an interaction with
a choose-your-own-adventure system consisting of \choices\ and \select\ if:

@@

As we will see when we revisit the earlier examples formally, this way of invoking a choose-your-own-adventure
system is used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
In those systems, the user triggers the completion on a proof (program) containing a hole.
They then fill the hole and, possibly iteratively, further holes in the generated proof. The final
expression is embedded in the source code, but it does not indicate what options, identified by
$\iota_1, \ldots, \iota_n$, were selected in the process.

\subparagraph{Reconstructible completion.}
In many choose-your-own-adventure systems, it is possible to reconstruct a trace of the interaction
through which a program was constructed. This is the case with type providers, where a user chooses
a sequence of object members to be accessed and those members directly appear in the source code.
The same would be the case in a completion system for Coq that would offer tactics to apply,
becuase the resulting proof would contain a record of the selected tactics.

For systems such as type providers, we say that that expression completion is \emph{reconstructible}.
To capture the notion formally, we again need the \ident{init} operation, but also an operation
\ident{decode} that extracts identifiers of invoked completions from an expression:

\begin{definition}[Reconstructible expression completion]
An expression $E[?]$ is reconstructibly completed as $E[e]$ through an interaction with
a choose-your-own-adventure system consisting of \choices\ and \select\ if:

\vspace{-0.5em}
\raggedright
\begin{enumerate}
\item $E[?]$ is completed as $E[e]$ using \ident{init}, \choices\ and \select\\
  through a series of choices designated by identifiers $\iota_1, \ldots, \iota_n$,
\item it also holds that $\ident{decode}(e)=(\iota_1, \ldots, \iota_n)$.
\end{enumerate}
\end{definition}

If the integration of a choose-your-own-adventure system with a host programming language
uses reconstructible expression completion, we can recover the choices through which the user completed
an expression $e$ in a completion context $E[e]$, assuming they used the interactive system rather
than entering the code directly. This also means that we can reconstruct the final state $\sigma_n$
of the system by starting from $\ident{init}(E[?])$ and following the choices specified by
$\iota_1, \ldots, \iota_n$.

@@

\newpage
\begin{figure}[t]
\small
(a) Grammar of a minimal language with declarations and expressions
\[
\begin{array}{rcl}
\mathit{expr} & := & \mathit{literal} ~~|~~ \mathit{expr}+\mathit{expr} ~~|~~ \mathit{ident} ~~|~~ \mathit{expr}(\mathit{expr}^{*})\\
\mathit{decl} & := & \kvd{function}~\mathit{ident}(\mathit{ident}^{*}) \{ \mathit{block} \} ~~|~~ \kvd{let}~\mathit{ident}=\mathit{expr}\\\
\mathit{stmt} & := & \mathit{expr} ~~ | ~~ \mathit{decl}\\
\mathit{block} & := & \mathit{stmt}^{*}
\end{array}
\]
\begin{minipage}[t]{.5\textwidth}
(b) Construction based on the grammar

\begin{enumerate}
\setlength{\itemsep}{0.75em}
\item Add $\mathit{decl}$ as the first $\mathit{stmt}$ and fill two\\
  $\mathit{ident}$ leafs (name and parameters)\\[0.5em]
  $\begin{array}{l}
  \kvd{function}~\ident{sayHello}(\ident{who})~\{\\
  \quad\mathit{stmt}^{*}\\
  \}\\
  \mathit{stmt}^{*}
  \end{array}$

\item Sequentiall fill holes with invocations\\[0.5em]
  $\begin{array}{l}
  \kvd{function}~\ident{sayHello}(\ident{who})~\{\\
  \quad\ident{print}(\texttt{''Hello ''} + \ident{who})\\
  \}\\
  \ident{sayHello}(\texttt{''world''})
  \end{array}$
\end{enumerate}
\end{minipage}%
\begin{minipage}[t]{.5\textwidth}
(c) Alternative sequence based on refactorings

\begin{enumerate}
\setlength{\itemsep}{0.75em}
\item Start with a small working program\\[0.5em]
  $\begin{array}{l}
  \ident{print}(\texttt{''Hello world''})
  \end{array}$

\item Extract literal into a variable\\[0.75em]
  $\begin{array}{l}
  \kvd{let}~\ident{who} = \texttt{''world''}\\
  \ident{print}(\texttt{''Hello ''} + \ident{who})
  \end{array}$

\item Extract function with a parameter\\[0.75em]
  $\begin{array}{l}
  \kvd{function}~\ident{sayHello}(\ident{who})~\{\\
  \quad\ident{print}(\texttt{''Hello ''} + \ident{who})\\
  \}\\
  \ident{sayHello}(\texttt{''world''})
  \end{array}$

\end{enumerate}
\end{minipage}
\caption{Comparison of a naive grammar-based program construction (b) and more user-friendly
  program construction through a sequence of refactorings (c).}
\label{fig:grammar}
\end{figure}

\subsection{Structure editors}
Structure editors form a broad range of systems that support different kinds of interactions
for constructing programs. This allows us to demonstrate some of the design considerations for
choose-your-own-adventure systems. We first look at a simple choose-your-own-adventure system
derived from a context-free grammar, which is inspired by Sandblocks~\cite{beckmann-2023-all}.
We then discuss how such system could support refactorings in a way similar to
\textsc{Deuce}~\cite{hempel-2018-deuce}, extending the basic grammar-directed way of working.

\subparagraph{Formal model.}
We model a system that, similarly to Sandblocks, generates choices automatically based on a given
context-free grammar. We assume the grammar $(\mathcal{N},\Sigma,\mathcal{P},S)$ consists of set of
non-terminal symbols $\mathcal{N}$, terminal symbols $\Sigma$, start symbol $S\in\mathcal{N}$
and production rules $\mathcal{P}$ of the form $n\mapsto\boldsymbol{s}$ where
$n\in\mathcal{N}$ and $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$ (we write $n$ for non-terminals,
$t$ for terminals and $s$ for all symbols; bold indicates a sequence of symbols).

When interacting with the choose-your-own-adventure system, the user is gradually constructing a
program (sentence) of a form $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$. The system state is the program (sentence)
constructed so far, i.e., $\sigma = \boldsymbol{s}$. The process can continue
as long as there are non-terminal symbols in the sentence and production rules applicable to them.

In general, the $\choices$ operation could recommend applicable production rules for any
non-terminal in the sentence. The following always chooses the first non-terminal by assuming
a sentence $\boldsymbol{t}n\boldsymbol{s}$ consisting of a sequence of terminals, followed by a
single non-terminal and a sequence of arbitrary symbols (we discuss sequentialization
in Section~\ref{sec:limitations}). The $\choices$ operation then finds all applicable production
rules and offers them as choices. In a real-world system, identifiers $\iota_i$ are formed by
the names of the rules (omitted from our formalism).
\[
\begin{array}{rcl}
\choices(\boldsymbol{t}n\boldsymbol{s}) &=& \{\; \iota_i \mapsto \boldsymbol{t}\boldsymbol{s_i}\boldsymbol{s} ~|~ \forall (n\mapsto\boldsymbol{s_i})\in \mathcal{P} ~\}\\
\select(\boldsymbol{s}) &=& \boldsymbol{s}
\end{array}
\]
Unlike in the case of data exploration, the $\select$ operation does not ensure
that a program is fully constructed and the returned sentence may contain further non-terminals.

\subparagraph{Program refactorings.}
The formal model derived from a context-free grammar is an adequate model for structure
editors such as Sandblocks~\cite{beckmann-2023-all}, but many structure editors offer more
advanced operations for program construction. \textsc{Deuce}~\cite{hempel-2018-deuce},
which is a structure editor for a simple programming environment for creating Scalable
Vector Graphics (SVG) images, for example, makes it possible to transform program through
a range of refactorings such as extract function, introduce a local variable, inline definition,
reorder arguments and more.

Figure~\ref{fig:grammar} illustrates program construction via refactoring using a simple
example. It defines a small language with a grammar. Figure~\ref{fig:grammar} (b) shows some
of the steps necessary to construct a simple program via the grammar rules.
Figure~\ref{fig:grammar} (c) shows a more appealing alternative where the programmer starts
with a simple program and then gradually makes it more general through a sequence of refactorings.

Modelling refactorings formally requires a more expressive framework~\cite{steimann-2011-refactoring},
so we omit the details. The $\choices$ operation would need to identify applicable refactorings
for the current program and offer them to the user.

An important point illustrated by the two examples is whether the expression completion
implemented by the two sketched choose-your-own-adventure systems is reconstructible.
For a system based solely on context-free grammar, this is the case if the grammar is not
ambiguous. Although the constructed expression does not directly contain an encoding of the
sequence of operations, it is possible to reconstruct which production rules have been applied
(by parsing the expression according to the grammar). A choose-your-own-adventure structure
editor that supports refactorings no longer has this property, because it always offers multiple
ways of constructing a given program (a program can be constructed with or without a refactoring).

@@

\item For AI assistants (Section~\ref{sec:examples-aias}), the system offers a set of constraints
  based on the current inferred program. It does not let the user construct arbitrary constraints.
  Moreover, because the $\mathit{arg\,max}$ operation used in $\select$ performs statistical
  optimization, there is no guarantee that it can be used to generate a specific program.
  The system is only complete if it is possible to choose a constraint set $\boldsymbol{c}$ that
  restrict the set of programs $E_{\boldsymbol{c}}$ to a single given program. This is the case
  only for some AI assistants (e.g., datadiff, which always offers constraints to map column to
  any chosen other column).

@@

\subparagraph{Uniqueness.}

@@

\section{Limitations}
\label{sec:limitations}

@@

\noindent
TODO: Something about unnecessary seqentialization
